Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-02 doc/types-pldi2013/langdef.tex
coherence

The proof, which can be found in the submitted appendix, is lengthy but
straightforward due to the compositional operational semantics.
%It proceeds by standard induction on the structure of the
%derivation in the operational semantics, at each step showing that
%subexpressions have consistent starting environments and that the
%value and trace consistency of the subexpressions is sufficient to prove
%the consistency of the parent expression.

@@

An important insight from the proof of Theorem~\ref{thm:effects} is that 
it does not rely on the definition of {\em valid\_interleave}.  Weakening 
{\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:opsemfns}), 
which only requires any arbitrary interleaving of memory
traces from subexpressions, 
would have been a sufficient condition for the proof.  
The additional constraints in {\em valid\_interleave}
(Figure~\ref{fig:validinterleave}) address the {\em coherence} of heap accesses.
Coherence defines the permissible ordering of operations to each individual heap 
location by parallel computations.  The constraints enforced by {\em valid\_interleave}
specify exactly what interleavings of memory traces are permitted based on
the per-region coherence modes introduced by Legion for supporting parallelism
in the presence of aliasing.

To determine whether an interleaving of two or more memory traces is valid, we consider three sets of addresses:
\begin{itemize}
\item exclusive locations ($l \in L_{excl}$) are those which have at least one access in exclusive mode in the traces
and are not in the clobber set.  For these locations, we require sequential execution semantics - all
reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
as if all writes and reductions were applied from each trace in order.
\item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
are in neither $L_{excl}$ nor the clobber set.  For these locations, we allow permutations of the original
subexpression trace order.
\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
enforced.  The validity of the interleaving of these accesses will be performed in the enclosing expression.
\end{itemize}

@@

Although it is our goal to enable parallel execution whenever possible, we briefly
consider sequential execution of a Legion program.  We show that a sequential execution trivially
satisfies the interleaving criteria required by the Legion operational semantics.  Our proof
of the soundness of parallel scheduling depends on this result.
%In addition to
%sequential execution being the default behavior of the hierarchical scheduler, our proof of the
%soundness of the parallel scheduling will depend on this result.

Sequential execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
value and memory trace that result from a sequential execution of an expression are always valid executions of
Legion programs.

@@

To determine when parallel execution is safe, we start from the sequential execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution and parallel execution is safe.  To
efficiently discover these cases at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
the current region mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.

We define a {\em non-interference} operator $\epsilon_1 \nonint \epsilon_2$ on two memory operations as follows:
$$
\begin{array}{l@{ }l}
op_1(l_1, c_1,& v_1)\ \nonint\ op_2(l_2, c_2, v_2) \Leftrightarrow l_1 \not= l_2 \vee \\
& (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2)
\end{array}
$$
Accesses to different locations cannot effect each other, and reads have no side-effects.  The safety of the third case
follows from the requirement that reduction operations be commutative.

The next step is to show that this non-interference operator is strong enough to preserve the validity of an interleaving
of two or more traces when reordering a pair of memory operations.
\begin{lem}
\label{lem:nonintswap}
\rm
Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
$\epsilon_1, \epsilon_2$ be two memory operations from $E_i$ and $E_j$ ($i \not= j$).  Then,
$$\begin{array}{l@{}l}
valid\_interleave&(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \Rightarrow \\
 & valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
\end{array}$$
\end{lem}
The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.

We now extend the notion of non-interference to whole memory traces, with two traces being non-interfering if no operation from one
trace interferes with any operation from the other:
$$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$

If whole memory traces are non-interfering, then any interleaving of them can be sorted via 
pairwise swaps to match the sequential memory
trace. This gives us the following result which permits safe parallel execution:

@@

We now show that we can turn this result into an efficient runtime test for non-interface that can be 
performed task execution.  The key to this test is the privileges on Legion tasks and their ability to
provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
extended to work on privileges and sets of privileges:

@@

The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have exact equivalents
for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
approximated by two tests.  The first uses (logical) region disjointness constraints from the type system to 
statically infer non-interference.  The second uses the region mapping $M$ to dynamically determine the
disjointness of the two regions.  Although still a dynamic test, it is performed
once per pair of regions rather than for every pair of memory operations.  Details on how to perform
the dynamic test efficiently at runtime can be found in \cite{Legion12}.
%and often need not be performed at all
%(if one of the other three conditions is shown to hold at compile-time).

As region non-interference is an approximation of memory trace non-interference, we must show that it is sound.  We
state the Lemma here, and give the proof in \cite{LegionTypes12}.

@@

%We are now able to put all the pieces together and 
We now state the theorem that allows the Legion runtime to perform hierarchical
and parallel scheduling of non-interfering tasks:

@@

The proof follows directly from Lemmas~\ref{lem:noninteffects}
and \ref{lem:nonintpriv}.  This result holds even if the
clobber set $C$ is non-empty, allowing two locally independent
subtasks to run in parallel even if one or both of them is interacting
(in a programmer-permitted way) with a third subtask.

We now highlight an important aspect of the Legion
implementation that is different from other systems and relies on the
soundness of privileges.  Dynamic non-interference of 
memory operations is something that can only be determined after 
evaluation of an expression is completed, and only at great expense, 
as illustrated by work on transactional
memory \cite{Harris05}.  However, many systems are willing to pay this cost to
extract parallelism from operations that access aliased data. At the other extreme are systems like 
Jade \cite{Rinard98} and DPJ \cite{Bocchino11} that check non-interference 
statically by disallowing aliasing.  In contrast, Legion checks non-interference of privileges
at runtime, which is much simpler and more efficient than checking
non-interference of dynamic memory traces.  Even though the privileges
themselves are static, the region mapping $M$ is dynamic.  Dynamically testing non-interference 
on the privileges of physical regions allows parallel execution in many more cases than a purely 
static analysis can achieve in the presence of aliasing.

@@

In cases where Legion cannot safely infer non-interference of
privileges, either because the programmer did not sufficiently partition the regions to expose disjointness or because the two
tasks actually access the same data in aliased regions, parallel execution can
still be requested by the programmer through the use of relaxed coherence 
annotations on individual regions passed to a task.
%the default behavior of sequential execution can
%be avoided through the use of relaxed coherence.

The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
tasks access interfere due to accessing aliased regions, they may execute in either order, allowing
the task issued later in program order to run as soon as its ready, even before the task issued earlier 
in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).

To show that this is safe, we define a relaxed version of non-interference for atomic coherence as
follows:
$$
\begin{array}{l@{}l}
op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee \\
& (c_1 = atomic \wedge c_2 = atomic)
\end{array}$$

@@

The other way in which coherence can be relaxed is using the {\em simult} mode, which allows
multiple tasks to access the same region concurrently.  The simult coherence mode is appropriate in two
important cases:
\begin{enumerate}
\item When subtasks are accessing disjoint data, but the disjointness is hard to describe (e.g. walking
separate linked lists that have been allocated in the same region)
\item When the algorithm is tolerant of non-determinism (e.g. in a breadth-first search, setting the parent
pointer of a node with multiple equally-short paths to the root).
\end{enumerate}

To support the {\em simult} coherence mode, the non-interference test is extended 
%When one or more regions have been marked as using the simult coherence mode, 
%the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
with a $\nonint^S$ operator analogous to the one for the atomic coherence case.
Because the rules for valid interleavings exclude locations that are only accessed in simult mode, it is straightforward
to extend Theorem~\ref{thm:parallelexec} to show that parallel execution is safe as long as $\Phi_i \nonint[M]^S \Phi_j$.

It is also possible to have both atomic and simult coherence modes in use at the same time for different regions in a task call.
In this case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and
Theorem~\ref{thm:atomicexec} is extended to allow arbitrary reordering (but not simultaneous execution) of subtasks when
$\Phi_i \nonint[M]^{AS} \Phi_j$.

>>>>

The proof, which can be found in the submitted appendix, is made
straightforward by the compositional operational semantics.
%It proceeds by standard induction on the structure of the
%derivation in the operational semantics, at each step showing that
%subexpressions have consistent starting environments and that the
%value and trace consistency of the subexpressions is sufficient to prove
%the consistency of the parent expression.

@@

An important insight from the proof of Theorem~\ref{thm:effects} is that 
it does not rely on the definition of {\em valid\_interleave}.  Weakening 
{\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:validinterleave}), 
which  allows arbitrary interleavings of memory traces from subexpressions, 
is sufficient for the proof of soundness.  
The stronger constraints in {\em valid\_interleave}
(Figure~\ref{fig:validinterleave}) address the coherence of heap accesses;
{\em valid\_interleave} specifies permitted interleavings of memory operations for
the particular coherence modes on logical regions.

To determine whether an interleaving of two or more memory traces is valid, we consider three sets of addresses:
\begin{itemize}
\item exclusive locations ($l \in L_{excl}$) are those which have at least one access in exclusive mode in the traces
and are not in the clobber set.  For these locations, we require sequential execution semantics - all
reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
as if all writes and reductions were applied from each trace in order.
\item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
are in neither $L_{excl}$ nor the clobber set.  For these locations, we allow permutations of the original
subexpression trace order.
\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
enforced.  The validity of the interleaving of these accesses is performed by some enclosing expression.
\end{itemize}

@@

%Although it is our goal to enable parallel execution whenever possible, we briefly
%consider sequential execution of a Legion program.  
We now show that a sequential execution trivially
satisfies the interleaving criteria required by the operational semantics.  Our proof
of the soundness of parallel scheduling depends on this result.
%In addition to
%sequential execution being the default behavior of the hierarchical scheduler, our proof of the
%soundness of the parallel scheduling will depend on this result.

Sequential execution ignores the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and interleaves traces by concatenating the subexpressions' traces in program order.  By ignoring
the coherence modes, the clobber set remains empty and the result of all {\em read} expressions is fully
determined.  The following lemma and theorem show that the
value and memory trace that result from a sequential execution are always valid executions.

@@

To determine when parallel execution is safe, we start from the sequential execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution and parallel execution is safe.  To
efficiently discover these cases at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
the current region mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.
We define a {\em non-interference} operator $\epsilon_1 \nonint \epsilon_2$ on two memory operations:
$$
\begin{array}{l@{ }l}
op_1(l_1, c_1,& v_1)\ \nonint\ op_2(l_2, c_2, v_2) \Leftrightarrow l_1 \not= l_2 \vee \\
& (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2)
\end{array}
$$
Accesses to different locations cannot affect each other, and reads have no side-effects.  The safety of the third case
follows from the requirement that reduction operations be commutative.  
We next show that this non-interference operator preserves the validity of an interleaving
of two or more traces when reordering a pair of memory operations.
\begin{lem}
\label{lem:nonintswap}
\rm
Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
$\epsilon_1, \epsilon_2$ be two memory operations from $E_i$ and $E_j$ ($i \not= j$).  Then,
$$\begin{array}{l@{}l}
valid\_interleave&(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \Rightarrow \\
 & valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
\end{array}$$
\end{lem}
%The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.

\noindent Two whole memory traces are non-interfering if no operation from one
trace interferes with any operation from the other:
$$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
If whole memory traces are non-interfering, any interleaving can be sorted via 
pairwise swaps to match the sequential memory
trace. This gives us a result permitting safe parallel execution:

@@

\noindent 
We now use the bounds that static privileges place on 
runtime accesses to give an efficient runtime test for non-interference.  We first extend the non-interference operator 
to work on privileges:

@@

\noindent The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have equivalents
for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
approximated by two tests.  The first uses (logical) region disjointness constraints from the type system to 
statically infer non-interference.  The second uses the region mapping $M$ to dynamically determine the
disjointness of the two regions.  Although a dynamic test, it is performed
once per pair of regions rather than for every pair of memory operations.  An algorithm
to perform the dynamic test efficiently is given in \cite{Legion12}.
%and often need not be performed at all
%(if one of the other three conditions is shown to hold at compile-time).
As region non-interference is an approximation of memory trace non-interference, we must show that it is sound.  
%We state the Lemma here, and give the proof in \cite{LegionTypes12}.

@@

%We are now able to put all the pieces together and 
We now state the theorem that allows the Legion runtime to perform hierarchical
and parallel scheduling of non-interfering tasks.

@@

%The proof follows directly from Lemmas~\ref{lem:noninteffects} and \ref{lem:nonintpriv}.  
\noindent This result holds even if the
clobber set $C$ is non-empty, allowing locally independent
subtasks to run in parallel even if they interact
(in a programmer-permitted way) with another  subtask.

We highlight an important aspect of a Legion
implementation that is different from other systems and relies on the
soundness of privileges.  Dynamic non-interference of 
memory operations can only be determined after 
evaluation of an expression is completed, and only at great expense, 
as illustrated by work on transactional
memory \cite{Harris05}.  
%However, many systems are willing to pay this cost to
%extract parallelism from operations that access aliased data. 
At the other extreme are systems like 
Jade \cite{Rinard98} and DPJ \cite{Bocchino11} that check non-interference 
statically by disallowing aliasing.  In contrast, Legion checks non-interference of privileges
at runtime, which is much simpler and more efficient than checking
non-interference of dynamic memory traces.  Even though the privileges
themselves are static, the region mapping $M$ is dynamic.  Dynamically testing non-interference 
on the privileges of physical regions allows parallel execution in many more cases than a purely 
static analysis can achieve in the presence of aliasing.

@@

In cases where Legion cannot safely infer non-interference of
privileges (perhaps because two tasks actually access the same data in aliased regions) 
parallel execution can
still be requested by the programmer through the use of relaxed coherence 
annotations on individual regions passed to a task.
%the default behavior of sequential execution can
%be avoided through the use of relaxed coherence.
The {\em atomic} coherence mode specifies that although two
tasks interfere due to accessing aliased regions, they may execute in either order, allowing
the task issued later in program order to possibly run before the task issued earlier 
in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).
To show this is safe, we define a relaxed version of non-interference for atomic coherence:
$$
\begin{array}{l@{}l}
op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee \\
& (c_1 = atomic \wedge c_2 = atomic)
\end{array}$$

@@

Coherence also can be relaxed using the {\em simult} mode, which allows
multiple tasks to access the same region concurrently.  The simult coherence mode is appropriate in two
important cases:
\begin{enumerate}
\item When subtasks are accessing disjoint data, but the disjointness is difficult to describe (e.g. walking
separate linked lists that have been allocated in the same region).
\item When the algorithm is tolerant of non-determinism (e.g. in a breadth-first search, setting the parent
pointer of a node with multiple equally-short paths to the root).
\end{enumerate}

\noindent To support the {\em simult} coherence, the non-interference test is extended 
%When one or more regions have been marked as using the simult coherence mode, 
%the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
with a $\nonint^S$ operator, analogous to $\nonint^A$ for atomic coherence.
Because the rules for valid interleavings exclude locations that are only accessed in simult mode, it is straightforward
to extend Theorem~\ref{thm:parallelexec} to show that parallel execution is safe as long as $\Phi_i \nonint[M]^S \Phi_j$.

It is also possible to have both atomic and simult coherence modes at the same time for different regions in a task call.
In this case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and
Theorem~\ref{thm:atomicexec} is extended to allow arbitrary reordering (but not simultaneous execution) of subtasks when
$\Phi_i \nonint[M]^{AS} \Phi_j$.
